---
title: "individual1"
author: "Matthew Eisner"
date: "9/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Part two: Exploration

Use `ggplot2` to create the plots, and `dplyr` to manipulate the data frame as much as possible. You may want to rename the columns of the data frame for your convenience using `rename` in `dplyr`.

```{r}
library(dplyr)
library(ggplot2)
dat <- readxl::read_xls('./GSS.xls')

colnames(dat) <- c('year', 'happiness', 'religion_raised', 'political_party', 'sex', 'school_years', 'marital_status', 'work_hrs', 'id_num', 'ballot')

dat <- dat %>% mutate(school_years = as.numeric(school_years))
dat <- na.omit(dat)
```

1. Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.

```{r}
dat %>% group_by(political_party) %>% summarize(avg_years=mean(school_years), .groups='drop') %>% arrange(avg_years)
```

2. Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

```{r}
educ <- dat %>% group_by(political_party) %>% filter(school_years > median(school_years))
str(educ)
```    

For each of the investigation below, create at least one numerical summary and one appropriate visualization. Write a paragraph to describe your findings and the variables you used.

3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

```{r}
dat1 <- mutate(dat, happy = ifelse(happiness == 'Very Happy' | happiness == 'Pretty happy', TRUE, FALSE))
ggplot(dat1, aes(x=marital_status, fill=happy)) + geom_bar(position='fill')
```

4. Does party affiliation affect the relationship you found in the last question?

```{r}
ggplot(dat1, aes(x=marital_status, fill=happy)) + geom_bar(position='fill') + facet_wrap(~political_party) + theme(axis.text.x=element_text(angle =- 90, vjust = 0.5))
```

5. How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

```{r}
ggplot(dat, aes(x=marital_status, y=school_years)) + geom_boxplot()
```

6. Explore two more interesting questions, and answer those questions using the GSS data.

```{r}
dat <- dat %>% mutate(work_hrs = as.numeric(work_hrs))
dat <- na.omit(dat)

ggplot(dat, aes(x=work_hrs, y=school_years)) + geom_point() + geom_smooth(method = "lm", se=FALSE)
```

The team’s exploration will be evaluated according to the general interest and the depth of the exploration.
Tips and Resources

* Decide a common version of the dataset to work on
* Take a look at the key trends to find interesting questions to explore. 
