---
title: "individual1"
author: "Matthew Eisner"
date: "9/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Part two: Exploration

Use `ggplot2` to create the plots, and `dplyr` to manipulate the data frame as much as possible. You may want to rename the columns of the data frame for your convenience using `rename` in `dplyr`.

```{r}
library(dplyr)
library(ggplot2)
dat <- readxl::read_xls('./GSS.xls')

colnames(dat) <- c('year', 'happiness', 'religion_raised', 'political_party', 'sex', 'school_years', 'marital_status', 'work_hrs', 'id_num', 'ballot')

dat <- dat %>% mutate(school_years = as.numeric(school_years))
dat <- na.omit(dat)
```

1. Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.

```{r}
dat %>% group_by(political_party) %>% summarize(avg_years=mean(school_years), .groups='drop') %>% arrange(avg_years)
```

For each of the investigation below, create at least one numerical summary and one appropriate visualization. Write a paragraph to describe your findings and the variables you used.

3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

```{r}
ggplot(dat, aes(x=marital_status, fill=happiness)) + geom_bar()
```

4. Does party affiliation affect the relationship you found in the last question?


5. How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?


6. Explore two more interesting questions, and answer those questions using the GSS data.



The team’s exploration will be evaluated according to the general interest and the depth of the exploration.
Tips and Resources

* Decide a common version of the dataset to work on
* Take a look at the key trends to find interesting questions to explore. 
