---
title: "DS 202 - Lab 3"
author: "Matthew Eisner and Jeffrey Kinard"
date: "9/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Outline

* Work in assigned pairs, as in Lab 2
* Create a repository (one per team) & give access to your teammate
* For a refreshment of git & github, review the procedures for collaboration for Lab 2.
* Explore what leads to happiness using data from the General Social Survey
* Create an R markdown file that contains all code to read the data, create the plots, and your answers.
* Teammates will review and revise each other’s work.
* Submit!

## Submission Requirements

* A team member should create a GitHub repository called ds202_lab3 which includes the group work.

* Each team member should create at least one commit with message “Review and Revision”, which include changes on the analysis of your partner’s.

* The team will collaborate on team.Rmd to answer all the questions, and also include your individual exploration in either individual1.Rmd or individual2.Rmd. **Only the report generated from `team.Rmd` will be graded.**

* Include the data you used extracted from the General Social Survey website.

* The correct folder structure is as below. The folder ds202_lab3 is a git repository.

    ds202_lab3
    |-- individual1.Rmd
    |-- individual2.Rmd
    |-- team.Rmd
    |-- *.html
    |-- Data files...

## Background

The General Social Survey (GSS) has been run by NORC since 1972 to keep track of current opinions across the United States.

There is a data explorer which let you to interactively explore and extract the data here https://gssdataexplorer.norc.org/

## Part zero: Exploring the GSS data

You can search for information you are interested in under in Keyword search box. Try “happiness”. If you do not find desired result, click the dropdown arrow, and select “All” to search in all available information and not just the variable names. We will focus on variables that have a long span of availability over the years. Discuss interesting findings in your study group.

We will be exploring variables related to

* **Happiness**
* **Family (marital status)**
* **Education**
* **Political affilation**
* Many other possible topics, such as Work (hours) and Belief – it is up to you

## Part one: Extracting and loading data

1. Go to https://gssdataexplorer.norc.org/
2. Click Search variable
3. Search “happy”; click add to cart; go back. Repeat for “marital status”, “educ”, “respondent sex”, “political party affiliation”, and other variables (you decide)
4. After you finished adding the variables, click Actions – Save to Project
5. Register an account
6. Click Default Project
7. Click Actions – Extract data
8.  a. Name your extract (e.g. extract1) – click NEXT;
    b. ADD ALL variables – click NEXT;
    c. click NEXT;
    d. **Check Excel Workbook (data + metadata), and leave all others unchecked;**
    e. CREATE EXTRACT
9. Wait for the extract to be ready, and then download an archive file.
10. Extract the archive file (ending with .tar.gz). Windows users: You may need 7-Zip
11. Load the Excel file into R, using

install.packages('readxl')
dat <- readxl::read_xls('PATH_TO_EXL/GSS.xls')

## Part two: Exploration

Use `ggplot2` to create the plots, and `dplyr` to manipulate the data frame as much as possible. You may want to rename the columns of the data frame for your convenience using `rename` in `dplyr`.

```{r}
library(dplyr)
library(ggplot2)
dat <- readxl::read_xls('./GSS.xls')

colnames(dat) <- c('year', 'happiness', 'religion_raised', 'political_party', 'sex', 'school_years', 'marital_status', 'work_hrs', 'id_num', 'ballot')

dat <- dat %>% mutate(school_years = as.numeric(school_years))
dat <- na.omit(dat)
```

1. Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.

Those who did not answer the question were the only group to have less than 12 years of education, which is equivalent to on average, not finishing high school. All other political party affiliations, on average, finished high school and had some length of college/university level education.

```{r}
dat %>% group_by(political_party) %>% summarize(avg_years=mean(school_years), .groups='drop') %>% arrange(avg_years)
```

2. Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

```{r}
educ <- dat %>% group_by(political_party) %>% filter(school_years > median(school_years))
str(educ)
```

For each of the investigation below, create at least one numerical summary and one appropriate visualization. Write a paragraph to describe your findings and the variables you used.

3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

The only groups that had over 50% of respondents answer that they were happy were those who identified as divorced, never married, separated, or widowed. Those who were married or did not answer were the only groups where less than 50% of responses were happy.

```{r}
dat1 <- mutate(dat, happy = ifelse(happiness == 'Very Happy' | happiness == 'Pretty happy', TRUE, FALSE))
ggplot(dat1, aes(x=marital_status, fill=happy)) + geom_bar(position='fill')
```

4. Does party affiliation affect the relationship you found in the last question?

```{r}
ggplot(dat1, aes(x=marital_status, fill=happy)) + geom_bar(position='fill') + facet_wrap(~political_party) + theme(axis.text.x=element_text(angle =- 90, vjust = 0.5))
```

5. How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

```{r}
ggplot(dat, aes(x=marital_status, y=school_years)) + geom_boxplot()
```

6. Explore two more interesting questions, and answer those questions using the GSS data.



The team’s exploration will be evaluated according to the general interest and the depth of the exploration.
Tips and Resources

* Decide a common version of the dataset to work on
* Take a look at the key trends to find interesting questions to explore. 
